### 2023-09-29 appendices/supplementary_modelagnostic.tex
Move table captions to top
Updates to citation commands and footnotes

A crude diagonal approximation can be computed by simply estimating the diagonal of $I({\theta})$ and setting all off-diagonal elements to zero. Such diagonal approximations have been used in machine learning for decades: for instance, \textcite[Section 3.12.2]{lecun_modeles_1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying". Much more complex approximations have been derived, although diagonal approximations have been consistently used (e.g. by \citealp{kirkpatrick_overcoming_2017}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \citep{kingma_adam_2015} or RMSProp \citep{tieleman_lecture_2012}. A good discussion on these issues is provided in Martens's \citeyearpar{martens_new_2020} recent review.

@@

We show here that it is possible to re-interpret this score as a MMD score with a certain Fisher kernel. 
The generative model induced on $z$ by linear disciminant analysis is a Gaussian mixture:
\begin{equation}
    p_{\pi, \mu , \Sigma }(z) = \sum_{k=1}^K \pi_k \mathcal{N}(z | \mu_k, \Sigma).
\end{equation}
If we want a powerful deep kernel, it seems somewhat natural to consider the Fisher kernel associated with this generative model. The most important part of this mixture model are arguably the class-specific means (indeed, the model has been trained to discriminate the classes as well as possible). Therefore, we will only include these means in the Fisher kernel, and look at
\begin{equation}
	\Phi_\textup{Fisher} (x) = I(\mu)^{-1/2} \nabla_\mu  \log p_{\pi, \mu , \Sigma }(z),
\end{equation}
assuming that $\pi$ and $\Sigma$ are fixed at their maximum likelihood estimates. Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \citep{perronnin_improving_2010}. Our idea is to re-use ideas introduced by this computer vision litterature. Under the assumption that the Gaussian clusters are well-separated, \textcite{tanaka_fisher_2013}, extending an earlier analysis of \textcite[Appendix A]{sanchez_image_2013}, showed that
\begin{equation}
	[\Phi_\textup{Fisher} (x)]_{\mu_k} \approx \sqrt{\frac{p(z|k)}{\pi_k }}  \Sigma^{-1/2} (z - \mu_k).
\end{equation}
Now, using the fact that the expected value of the score is approximatively zero, we can write that
\begin{equation}
    \textup{MMD}_{\Phi_\textup{Fisher}}^2 \approx \sum_{k=1}^K  \parallel [\Phi_\textup{Fisher} (x)]_{\mu_k} ||_2^2 \approx \sum_{k=1}^K \frac{p(z|k)}{\pi_k} (z - \mu_k)^T \Sigma^{-1} (z - \mu_k).
\end{equation}

@@

\subsection{A bit more background}
The three considered DGMs are both parametrized by neural networks but they differ in the way they model the data distribution of interest. Assume we are interested in approximating a target distribution $p^*(\mathbf{x})$, for example a distribution of natural images, as it is done when using CIFAR10. PixelCNN++ is an autoregressive model and it models $p^*(\mathbf{x})$ as a product of conditional distribution over the variables, i.e.\@ $p(\mathbf{x}) = p(x_1) \prod_{d=2}^D p(x_d \mid \mathbf{x}_{<d})$, where $\mathbf{x}_{<d} = [x_1, \dots, x_{d-1}]^T$. Glow is a normalizing flow model and it approximate $p^*(\mathbf{x})$ by using a sequence of bijiective transformations starting from a simple distribution, also called base distribution. If we use only a single invertible transformation $f$, the normalizing  flow is defined as $\mathbf{x} = f(\mathbf{z})$, where $\mathbf{z} \sim p_{Z}(\mathbf{z})$, and $p_{X}(\mathbf{x}) = p_{Z}(\mathbf{z})|\det J_f(\mathbf{z}) |^{-1}$, where we used the change of variable formula. For these two types of model we have a tractable likelihood that can be used to optimize the model parameters. The Variational Autoencoder (VAE), instead, is a framework to model the data with a latent variable model, i.e.\@ $p(\mathbf{x}, \mathbf{z}) = p(\mathbf{x} \mid \mathbf{z}) p(\mathbf{z})$, where $\mathbf{x}$ is the observed input data and $\mathbf{z}$ is a stochastic latent variable and the prior distribution $p(\mathbf{z})$ is usually a standard Normal. Since the posterior $p(\mathbf{z}\mid\mathbf{x})$ is not tractable, a variational distribution $q_{\phi}(\mathbf{z}\mid\mathbf{x})$ is used as an approximation. Due to the intractability of the posterior, we cannot directly optimize the likelihood of the model, but instead the model parameters are optimized by maximizing the evidence lower bound (ELBO): $\log p_{\theta}(\mathbf{x}) \geq \mathbb{E}_{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \left[ \log \frac{p(\mathbf{x}, \mathbf{z})}{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \right] \equiv \mathcal{L} $. In this work we are considering an Hierarchical VAE (HVAE) with bottom-up inference as done in \cite{havtorn_hierarchical_2021}. This is an extension of the VAE framework that consider an hierarchy of $L$ latent variables $\mathbf{z} = \mathbf{z}_1,\dots, \mathbf{z}_L$. The bottom-up inference is defined as $q_{\phi}(\mathbf{z}\mid \mathbf{x}) = q_{\phi}(\mathbf{z}_1\mid \mathbf{x})\prod_{i=2}^L q_{\theta}(\mathbf{z}_i \mid \mathbf{z}_{i-1})$, while the generative path is top-down, meaning $p_{\theta}(\mathbf{x}\mid \mathbf{z}) = p(\mathbf{x}\mid \mathbf{z}_1)p_{\theta}(\mathbf{z}_1\mid \mathbf{z}_2)\cdots p_{\theta}(\mathbf{z}_{L-1}\mid \mathbf{z}_L)$. This is still trained by maximizing the ELBO. For a more in-depth explanation of these models we refer to their papers.

@@

\paragraph{Glow} For training Glow models we follow \textcite{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}. They closely follow \textcite{nalisnick_deep_2019} and \cite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers. While \textcite{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32. For the greyscale dataset our Glow is made up of 2 scales with 16 coupling layers, and a 3-layers highway network with 200 hidden units is used to predict the scale and shift parameters. For CIFAR10 and SVHN, instead, we used 3 scales with 8 coupling layers, and 400 hidden units for the 3-layers highway network. %Since training flow-based models is a bit unstable, in some cases we had to restart the training when the training failed. 
For a more in-depth description, we refer to the codebase and the Appendix C of \textcite{kirichenko_why_2020}.

\paragraph{Hierarchical VAE} We follow \cite{havtorn_hierarchical_2021} for both model architecture design and training choices for our hierarchical VAEs. We used their open-sourced repository\footnote{\href{https://github.com/JakobHavtorn/hvae-oodd}{\texttt{https://github.com/JakobHavtorn/hvae-oodd}}}. As mentioned in the paper, the HVAE model we used has a bottom-up inference path and a top-down generative path. We trained each model for 1000 epochs using Adam optimizer with learning rate $3e-4$ and a batch-size of $128$. All models were initialized using the data-dependent initialization and they used weight-normalization \citep{salimans_weight_2016}. In addition to that, we always consider a hierarchy of three latent variables. For greyscale images (MNIST and FashionMNIST) we used a latent dimension of $8-16-8$ for $\mathbf{z}_1, \mathbf{z}_2, \mathbf{z}_3$ respectively, while for natural images (CIFAR10 and SVHN) we used $8-16-32$. For a more in depth description of the model, we refer to Appendix B of \textcite{havtorn_hierarchical_2021}.

@@

\subsection{Harmonic Mean}
\label{appendix_modelagnostic:harmonic}
In the paper we mentioned that another way to combine $p$-values from different test statistics is the Harmonic mean \citep{wilson_harmonic_2019}. This is defined as:
\begin{equation}
    \mathring{p} = \frac{\sum_{i=1}^{k} w_i}{\sum_{i=1}^{k} w_i / p_i},
\end{equation}
where $w_1,\dots,w_k$ are weights that sum up to 1. In our setting, we considered equal weights, i.e.\@ $w_i = 1 / k$. Therefore, if we simply consider two test statistics $T_1$ and $T_2$ and corresponding $p$-values $p_1$ and $p_2$, the harmonic mean $p$-values becomes:
\begin{equation}
    \mathring{p} = \frac{2 p_1 p_2}{p_1 + p_2}.
\end{equation}

@@

            \textsc{PixelCNN++} (model1)  & 0.1553 & 0.8006 & 0.6406 & 0.8126 & 0.6457 & 0.6407 \\
            \textsc{PixelCNN++} (model2) & 0.1567 & 0.7923 & 0.7070 & 0.7955 & 0.6498 &  0.7067\\
            \textsc{Glow} (RMSProp)  & 0.0630 & 0.8585 & 0.7929 & 0.8621 & 0.8651 &  0.7940\\
            \textsc{Glow} (Adam)   & 0.0627 & 0.7844 & 0.7620 &  0.7838 &  0.8624 &  0.7655 \\
            \textsc{HVAE}  & 0.0455 & 0.8041 &  0.7268 & 0.7634 & 0.8845 & 0.7334\\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:mmd}
    }
    \vspace*{-0.85\baselineskip}
\end{table*}

@@

            \textsc{PixelCNN++} (model1)  & 0.6826 & 0.6667 & 0.6571  \\
            \textsc{PixelCNN++} (model2) & 0.7300 & 0.7105 & 0.7243 \\
            \textsc{Glow} (RMSProp)  &  0.8683 & 0.8551 & 0.8510\\
            \textsc{Glow} (Adam) & 0.8613 & 0.8493 & 0.8588  \\
            \textsc{HVAE}  & 0.8699 &  0.8525 & 0.8245 \\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:all_combinations}
    \vspace*{-\baselineskip}
\end{table*}

@@

\begin{table*}[tb]
    \centering
    %\renewcommand{\thetable}{4}
    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different Glow models on single and combined statistics.]{AUROC$\uparrow$ for single-sample OOD detection. In this table we are comparing two different Glow models trained on CIFAR10 by considering two different checkpoints with almost the same test log-likelihood. We can see that both statistics vary a bit.}
        \scriptsize
        \begin{tabular}{ccccc}
            \toprule
            &\multicolumn{4}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-5}
            & \multicolumn{2}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-3}\cmidrule(l){4-5}
            \textsc{models}  &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{Glow} (RMSProp) \{\emph{check1}\} & 0.8651 &  0.7940  &  0.8683 &  0.8510\\
            \textsc{Glow} (RMSProp) \{\emph{check2}\} & 0.8532 &  0.6894  & 0.8275 & 0.7815\\
            \midrule
            \textsc{Glow} (Adam)  \{\emph{check1}\} & 0.8624 &  0.7655 & 0.8613 & 0.8588\\
            \textsc{Glow} (Adam) \{\emph{check2}\} & 0.8558 & 0.7327 & 0.8402 & 0.8303\\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:variability}
    \vspace*{-\baselineskip}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.5]{graphics/paper_modelagnostic/cifar_hvae_type1_correct_font.pdf}
    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values.]{Type I and Type II errors versus the significance level $\alpha$ on the combination values. We can control the FDR only for $\alpha > 0.2$ in this case. For $\alpha >0.2$, since we are using Benjamini-Hochberg procedure, we get that the Type I error stays below identity line.}
    \label{fig_modelagnostic:type1_cifar}
    \vspace*{-\baselineskip}
\end{figure}

@@

\begin{table*}[tb]
    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing single and combined statistics using a Gaussian mixture model (GMM) trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN).]{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM). For Fisher's method we mean the combination of the typicality test and the test statistic. These are also combined using DoSE.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}  & 0.6627 & 0.5514  & 0.5196 & 0.8777  & 0.7689 & 0.8152 \\
            \textsc{100}   &  0.6872 & 0.5509 & 0.5575 & 0.8742 & 0.7965 & 0.7989 \\
            \bottomrule
            &  &  &   &   &  & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}   & 0.2335  &   0.6087 & 0.6759 & 0.3512 &  0.6098 & 0.6569 \\
            \textsc{100} & 0.2372 &  0.6136 & 0.6714 & 0.3294  &  0.5898  & 0.6573 \\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:gmm_results}
    }
    \vspace*{-\baselineskip}
\end{table*}

@@

\section{Yes, we should talk about CelebA}
\label{appendix_modelagnostic:celeba}
Out-of-distribution detection performance is not only influenced by the model architecture or the training process. Indeed, transformations applied to the input data play an important role by transforming a difficult task into an easier problem where the likelihood can detect OOD data. By looking at the different results for Glow trained on CIFAR10 and tested on CelebA shown in \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, \cite{morningstar_density_2021}, and \cite{ahmadian_likelihoodfree_2021} we can see that the AUROC scores obtain by the plain log-likelihood are pretty different. In \cite{hendrycks_deep_2019} and  \cite{kirichenko_why_2020} the log-likelihood gets a poor performance, confirming that CIFAR10-CelebA is a challenging pair for DGMs, while in \cite{morningstar_density_2021} the likelihood is able to distinguish OOD data. While the main reason for these different results can be due to model implementation and training procedure, we decided to investigate how different transformations can influence OOD detection. Indeed, CelebA examples originally have a shape of $(218, 178, 3)$ and to transform them into $(32, 32, 3)$-shaped images, as CIFAR10, we have to resize them and then crop their center. The resize function is performing an interpolation, therefore we analyze how different interpolation strategies influence the OOD task.

We considered three different interpolations: bilinear (default in PyTorch), Lanczos, and nearest. As can be seen from \cref{fig_modelagnostic:celeba}, these transformations mostly affect the sharpness of the images. In \cref{table:celeba} we show how the OOD performance changes for our considered models when testing on CelebA where we applied different interpolations. We can notice that when using the bilinear interpolation we get results that are pretty similar to \cite{hendrycks_deep_2019}, \cite{kirichenko_why_2020}, and \cite{ahmadian_likelihoodfree_2021} in terms of likelihood OOD performance. When using the nearest interpolation, instead, we get results that are closer to \cite{morningstar_density_2021}.

@@

    \vspace*{-\baselineskip}

>>>>

A crude diagonal approximation can be computed by simply estimating the diagonal of $I({\theta})$ and setting all off-diagonal elements to zero. Such diagonal approximations have been used in machine learning for decades: for instance, \textcite[Section 3.12.2]{lecun_modeles_1987} used a similar approximation of the Hessian matrix, and called it ``outrageously simplifying". Much more complex approximations have been derived, although diagonal approximations have been consistently used (e.g. by \citealp{kirkpatrick_overcoming_2017}, who used essentially the same approximation in a supervised context), and are linked to several adaptive optimisation techniques like Adam \parencite{kingma_adam_2015} or RMSProp \parencite{tieleman_lecture_2012}. A good discussion on these issues is provided in Martens's \citeyearpar{martens_new_2020} recent review.

@@

We show here that it is possible to re-interpret this score as a MMD score with a certain Fisher kernel. 
The generative model induced on $z$ by linear disciminant analysis is a Gaussian mixture:
\begin{equation}
    p_{\pi, \mu , \Sigma }(z) = \sum_{k=1}^K \pi_k \mathcal{N}(z | \mu_k, \Sigma).
\end{equation}
If we want a powerful deep kernel, it seems somewhat natural to consider the Fisher kernel associated with this generative model. The most important part of this mixture model are arguably the class-specific means (indeed, the model has been trained to discriminate the classes as well as possible). Therefore, we will only include these means in the Fisher kernel, and look at
\begin{equation}
	\Phi_\textup{Fisher} (x) = I(\mu)^{-1/2} \nabla_\mu  \log p_{\pi, \mu , \Sigma }(z),
\end{equation}
assuming that $\pi$ and $\Sigma$ are fixed at their maximum likelihood estimates. Similar mixture-based Fisher kernels have been very popular in the past, and were actually a key element of state-of-the art classification models on Imagenet before deep nets won the competition \parencite{perronnin_improving_2010}. Our idea is to re-use ideas introduced by this computer vision litterature. Under the assumption that the Gaussian clusters are well-separated, \textcite{tanaka_fisher_2013}, extending an earlier analysis of \textcite[Appendix A]{sanchez_image_2013}, showed that
\begin{equation}
	[\Phi_\textup{Fisher} (x)]_{\mu_k} \approx \sqrt{\frac{p(z|k)}{\pi_k }}  \Sigma^{-1/2} (z - \mu_k).
\end{equation}
Now, using the fact that the expected value of the score is approximatively zero, we can write that
\begin{equation}
    \textup{MMD}_{\Phi_\textup{Fisher}}^2 \approx \sum_{k=1}^K  \parallel [\Phi_\textup{Fisher} (x)]_{\mu_k} ||_2^2 \approx \sum_{k=1}^K \frac{p(z|k)}{\pi_k} (z - \mu_k)^T \Sigma^{-1} (z - \mu_k).
\end{equation}

@@

\subsection{A bit more background}
The three considered DGMs are both parametrized by neural networks but they differ in the way they model the data distribution of interest. Assume we are interested in approximating a target distribution $p^*(\mathbf{x})$, for example a distribution of natural images, as it is done when using CIFAR10. PixelCNN++ is an autoregressive model and it models $p^*(\mathbf{x})$ as a product of conditional distribution over the variables, i.e.\@ $p(\mathbf{x}) = p(x_1) \prod_{d=2}^D p(x_d \mid \mathbf{x}_{<d})$, where $\mathbf{x}_{<d} = [x_1, \dots, x_{d-1}]^T$. Glow is a normalizing flow model and it approximate $p^*(\mathbf{x})$ by using a sequence of bijiective transformations starting from a simple distribution, also called base distribution. If we use only a single invertible transformation $f$, the normalizing  flow is defined as $\mathbf{x} = f(\mathbf{z})$, where $\mathbf{z} \sim p_{Z}(\mathbf{z})$, and $p_{X}(\mathbf{x}) = p_{Z}(\mathbf{z})|\det J_f(\mathbf{z}) |^{-1}$, where we used the change of variable formula. For these two types of model we have a tractable likelihood that can be used to optimize the model parameters. The Variational Autoencoder (VAE), instead, is a framework to model the data with a latent variable model, i.e.\@ $p(\mathbf{x}, \mathbf{z}) = p(\mathbf{x} \mid \mathbf{z}) p(\mathbf{z})$, where $\mathbf{x}$ is the observed input data and $\mathbf{z}$ is a stochastic latent variable and the prior distribution $p(\mathbf{z})$ is usually a standard Normal. Since the posterior $p(\mathbf{z}\mid\mathbf{x})$ is not tractable, a variational distribution $q_{\phi}(\mathbf{z}\mid\mathbf{x})$ is used as an approximation. Due to the intractability of the posterior, we cannot directly optimize the likelihood of the model, but instead the model parameters are optimized by maximizing the evidence lower bound (ELBO): $\log p_{\theta}(\mathbf{x}) \geq \mathbb{E}_{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \left[ \log \frac{p(\mathbf{x}, \mathbf{z})}{q_{\phi}(\mathbf{z}\mid \mathbf{x})} \right] \equiv \mathcal{L} $. In this work we are considering an Hierarchical VAE (HVAE) with bottom-up inference as done in \parencite{havtorn_hierarchical_2021}. This is an extension of the VAE framework that consider an hierarchy of $L$ latent variables $\mathbf{z} = \mathbf{z}_1,\dots, \mathbf{z}_L$. The bottom-up inference is defined as $q_{\phi}(\mathbf{z}\mid \mathbf{x}) = q_{\phi}(\mathbf{z}_1\mid \mathbf{x})\prod_{i=2}^L q_{\theta}(\mathbf{z}_i \mid \mathbf{z}_{i-1})$, while the generative path is top-down, meaning $p_{\theta}(\mathbf{x}\mid \mathbf{z}) = p(\mathbf{x}\mid \mathbf{z}_1)p_{\theta}(\mathbf{z}_1\mid \mathbf{z}_2)\cdots p_{\theta}(\mathbf{z}_{L-1}\mid \mathbf{z}_L)$. This is still trained by maximizing the ELBO. For a more in-depth explanation of these models we refer to their papers.

@@

\paragraph{Glow} For training Glow models we follow \textcite{kirichenko_why_2020} and their repository\footnote{\href{https://github.com/PolinaKirichenko/flows_ood}{\texttt{https://github.com/PolinaKirichenko/flows\_ood}}}. They closely follow \textcite{nalisnick_deep_2019} and \parencite{kingma_glow_2018} implementation for multi-scale Glow, where a scale is defined as the sequence of actorm, invertible $1\times1$ convolution and coupling layers. While \textcite{kirichenko_why_2020} only considers the RMSProp optimizer, we trained two different models, one using RMSProp and one using Adam with batch-size 32. For the greyscale dataset our Glow is made up of 2 scales with 16 coupling layers, and a 3-layers highway network with 200 hidden units is used to predict the scale and shift parameters. For CIFAR10 and SVHN, instead, we used 3 scales with 8 coupling layers, and 400 hidden units for the 3-layers highway network. %Since training flow-based models is a bit unstable, in some cases we had to restart the training when the training failed. 
For a more in-depth description, we refer to the codebase and the Appendix C of \textcite{kirichenko_why_2020}.

\paragraph{Hierarchical VAE} We follow \parencite{havtorn_hierarchical_2021} for both model architecture design and training choices for our hierarchical VAEs. We used their open-sourced repository\footnote{\href{https://github.com/JakobHavtorn/hvae-oodd}{\texttt{https://github.com/JakobHavtorn/hvae-oodd}}}. As mentioned in the paper, the HVAE model we used has a bottom-up inference path and a top-down generative path. We trained each model for 1000 epochs using Adam optimizer with learning rate $3e-4$ and a batch-size of $128$. All models were initialized using the data-dependent initialization and they used weight-normalization \parencite{salimans_weight_2016}. In addition to that, we always consider a hierarchy of three latent variables. For greyscale images (MNIST and FashionMNIST) we used a latent dimension of $8-16-8$ for $\mathbf{z}_1, \mathbf{z}_2, \mathbf{z}_3$ respectively, while for natural images (CIFAR10 and SVHN) we used $8-16-32$. For a more in depth description of the model, we refer to Appendix B of \textcite{havtorn_hierarchical_2021}.

@@

\subsection{Harmonic Mean}
\label{appendix_modelagnostic:harmonic}
In the paper we mentioned that another way to combine $p$-values from different test statistics is the Harmonic mean \parencite{wilson_harmonic_2019}. This is defined as:
\begin{equation}
    \mathring{p} = \frac{\sum_{i=1}^{k} w_i}{\sum_{i=1}^{k} w_i / p_i},
\end{equation}
where $w_1,\dots,w_k$ are weights that sum up to 1. In our setting, we considered equal weights, i.e.\@ $w_i = 1 / k$. Therefore, if we simply consider two test statistics $T_1$ and $T_2$ and corresponding $p$-values $p_1$ and $p_2$, the harmonic mean $p$-values becomes:
\begin{equation}
    \mathring{p} = \frac{2 p_1 p_2}{p_1 + p_2}.
\end{equation}

@@

            \textsc{PixelCNN++} (model1)  & 0.1553 & 0.8006 & 0.6406 & 0.8126 & 0.6457 & 0.6407 \\
            \textsc{PixelCNN++} (model2) & 0.1567 & 0.7923 & 0.7070 & 0.7955 & 0.6498 &  0.7067\\
            \textsc{Glow} (RMSProp)  & 0.0630 & 0.8585 & 0.7929 & 0.8621 & 0.8651 &  0.7940\\
            \textsc{Glow} (Adam)   & 0.0627 & 0.7844 & 0.7620 &  0.7838 &  0.8624 &  0.7655 \\
            \textsc{HVAE}  & 0.0455 & 0.8041 &  0.7268 & 0.7634 & 0.8845 & 0.7334\\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:mmd}
    }
    % \vspace*{-0.85\baselineskip}
\end{table*}

@@

            \textsc{PixelCNN++} (model1)  & 0.6826 & 0.6667 & 0.6571  \\
            \textsc{PixelCNN++} (model2) & 0.7300 & 0.7105 & 0.7243 \\
            \textsc{Glow} (RMSProp)  &  0.8683 & 0.8551 & 0.8510\\
            \textsc{Glow} (Adam) & 0.8613 & 0.8493 & 0.8588  \\
            \textsc{HVAE}  & 0.8699 &  0.8525 & 0.8245 \\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:all_combinations}
    % \vspace*{-\baselineskip}
\end{table*}

@@

\begin{table*}[tb]
    \centering
    %\renewcommand{\thetable}{4}
    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing two different Glow models on single and combined statistics.]{AUROC$\uparrow$ for single-sample OOD detection. In this table we are comparing two different Glow models trained on CIFAR10 by considering two different checkpoints with almost the same test log-likelihood. We can see that both statistics vary a bit.}
        \scriptsize
        \begin{tabular}{ccccc}
            \toprule
            &\multicolumn{4}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-5}
            & \multicolumn{2}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-3}\cmidrule(l){4-5}
            \textsc{models}  &  \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{Glow} (RMSProp) \{\emph{check1}\} & 0.8651 &  0.7940  &  0.8683 &  0.8510\\
            \textsc{Glow} (RMSProp) \{\emph{check2}\} & 0.8532 &  0.6894  & 0.8275 & 0.7815\\
            \midrule
            \textsc{Glow} (Adam)  \{\emph{check1}\} & 0.8624 &  0.7655 & 0.8613 & 0.8588\\
            \textsc{Glow} (Adam) \{\emph{check2}\} & 0.8558 & 0.7327 & 0.8402 & 0.8303\\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:variability}
    % \vspace*{-\baselineskip}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.5]{graphics/paper_modelagnostic/cifar_hvae_type1_correct_font.pdf}
    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values.]{Type I and Type II errors versus the significance level $\alpha$ on the combination values. We can control the FDR only for $\alpha > 0.2$ in this case. For $\alpha >0.2$, since we are using Benjamini-Hochberg procedure, we get that the Type I error stays below identity line.}
    \label{fig_modelagnostic:type1_cifar}
    % \vspace*{-\baselineskip}
\end{figure}

@@

\begin{table*}[tb]
    \caption[AUROC$\uparrow$ for single-sample OOD detection comparing single and combined statistics using a Gaussian mixture model (GMM) trained on FashionMNIST (tested on MNIST) and CIFAR10 (tested on SVHN).]{AUROC$\uparrow$ for single-sample OOD detection using a Gaussian mixture model (GMM). For Fisher's method we mean the combination of the typicality test and the test statistic. These are also combined using DoSE.}
    \resizebox{\textwidth}{!}{
        \scriptsize
        \begin{tabular}{ccccccc}
            \toprule
            &\multicolumn{6}{c}{\textsc{FashionMNIST (in) / MNIST (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}  & 0.6627 & 0.5514  & 0.5196 & 0.8777  & 0.7689 & 0.8152 \\
            \textsc{100}   &  0.6872 & 0.5509 & 0.5575 & 0.8742 & 0.7965 & 0.7989 \\
            \bottomrule
            &  &  &   &   &  & \\
            \toprule
            &\multicolumn{6}{c}{\textsc{CIFAR10 (in) / SVHN (out)}}\\
            \cmidrule{2-7}
            & \multicolumn{4}{c}{\textcolor{blue}{\textsc{single statistics}}} & \multicolumn{2}{c}{\textcolor{red}{\textsc{combination}}}\\
            \cmidrule(r){2-5}\cmidrule(l){6-7}
            \textsc{components}  & \textcolor{blue}{\textsc{$\log p(x)$}} & \textcolor{blue}{\textsc{$\|\nabla \log p(x)\|_2$}} & \textcolor{blue}{\textsc{Typicality}} & \textcolor{blue}{\textsc{Score stat}} & \textcolor{red}{\textsc{Fisher's method}} & \textcolor{red}{\textsc{DoSE$_{\textup{KDE}}$}} \\
            \midrule
            \textsc{50}   & 0.2335  &   0.6087 & 0.6759 & 0.3512 &  0.6098 & 0.6569 \\
            \textsc{100} & 0.2372 &  0.6136 & 0.6714 & 0.3294  &  0.5898  & 0.6573 \\
            \bottomrule
        \end{tabular}
        \label{tab_modelagnostic:gmm_results}
    }
    % \vspace*{-\baselineskip}
\end{table*}

@@

\section{Yes, we should talk about CelebA}
\label{appendix_modelagnostic:celeba}
Out-of-distribution detection performance is not only influenced by the model architecture or the training process. Indeed, transformations applied to the input data play an important role by transforming a difficult task into an easier problem where the likelihood can detect OOD data. By looking at the different results for Glow trained on CIFAR10 and tested on CelebA shown in \parencite{hendrycks_deep_2019}, \parencite{kirichenko_why_2020}, \parencite{morningstar_density_2021}, and \parencite{ahmadian_likelihoodfree_2021} we can see that the AUROC scores obtain by the plain log-likelihood are pretty different. In \parencite{hendrycks_deep_2019} and  \parencite{kirichenko_why_2020} the log-likelihood gets a poor performance, confirming that CIFAR10-CelebA is a challenging pair for DGMs, while in \parencite{morningstar_density_2021} the likelihood is able to distinguish OOD data. While the main reason for these different results can be due to model implementation and training procedure, we decided to investigate how different transformations can influence OOD detection. Indeed, CelebA examples originally have a shape of $(218, 178, 3)$ and to transform them into $(32, 32, 3)$-shaped images, as CIFAR10, we have to resize them and then crop their center. The resize function is performing an interpolation, therefore we analyze how different interpolation strategies influence the OOD task.

We considered three different interpolations: bilinear (default in PyTorch), Lanczos, and nearest. As can be seen from \cref{fig_modelagnostic:celeba}, these transformations mostly affect the sharpness of the images. In \cref{table:celeba} we show how the OOD performance changes for our considered models when testing on CelebA where we applied different interpolations. We can notice that when using the bilinear interpolation we get results that are pretty similar to \parencite{hendrycks_deep_2019}, \parencite{kirichenko_why_2020}, and \parencite{ahmadian_likelihoodfree_2021} in terms of likelihood OOD performance. When using the nearest interpolation, instead, we get results that are closer to \parencite{morningstar_density_2021}.

@@

    % \vspace*{-\baselineskip}
